::::{#exm-antiderivative-power}
#### Antiderivative of a power function

For $f(x) = x^2$, an antiderivative is $F(x) = \frac{x^3}{3}$,
since $\deriv{x}\frac{x^3}{3} = x^2 = f(x)$.

Adding any constant $C$ gives another antiderivative;
for example, with $C = 7$, $F(x) = \frac{x^3}{3} + 7$ also satisfies
$F'(x) = x^2$, since adding a constant does not change the derivative.
See @fig-antiderivatives.

{{< slidebreak >}}

```{r}
#| label: fig-antiderivatives-setup
#| include: false
C_vals <- c(-2, -1, 0, 1, 2)
x_lim <- c(-2, 2)
```

:::{#fig-antiderivatives layout-ncol="2"}

:::{#fig-antiderivatives-f}

```{r}
#| label: fig-antiderivatives-f-code
#| code-fold: true
ggplot() +
  geom_function(fun = \(x) x^2, xlim = x_lim, linewidth = 1) +
  labs(x = "x", y = expression(f(x))) +
  theme_minimal()
```

The function $f(x) = x^2$.
:::

:::{#fig-antiderivatives-F}

```{r}
#| label: fig-antiderivatives-F-code
#| code-fold: true
x_seq <- seq(x_lim[1], x_lim[2], length.out = 200)
df <- do.call(rbind, lapply(C_vals, \(C) {
  data.frame(
    x = x_seq,
    y = x_seq^3 / 3 + C,
    C = factor(C)
  )
}))

ggplot(df, aes(x = x, y = y, color = C)) +
  geom_line(linewidth = 0.8) +
  labs(x = "x", y = expression(F(x)), color = "C") +
  theme_minimal()
```

Family of antiderivatives $F(x) = x^3/3 + C$.
:::

The function $f(x) = x^2$ and five antiderivatives $F(x) = x^3/3 + C$ for $C \in \{`r paste(C_vals, collapse = ", ")`\}$.
Each antiderivative has the same derivative $f$; they differ only by a vertical shift.

:::

::::
