{{< slidebreak >}}

:::{#exm-stationary-least-squares}
#### A stationary point of the sum of squared errors, on real data

The Auto data record the horsepower and fuel economy (miles per gallon)
of `{r} nrow(read.csv("data/auto.csv"))` car models
sold between 1970 and 1982.
They come from the StatLib library at Carnegie Mellon University
and ship with the book *An Introduction to Statistical Learning*,
in the Python package [`ISLP`](https://islp.readthedocs.io/)
(`ISLP.load_data("Auto")`)
and the R package [`ISLR2`](https://cran.r-project.org/package=ISLR2)
(`ISLR2::Auto`).
This repository holds a copy at `data/auto.csv`,
and both code tabs below read that file.

Let $y_i$ be the miles per gallon and $x_i$ the horsepower of car $i$.
To predict miles per gallon from horsepower with a line,
write $\vbeta = \tp{(\beta_0, \beta_1)}$ and let $\mX$ be the matrix
whose row $i$ is $(1, x_i)$.
The sum of squared errors is

$$
f(\vbeta) = \norm{\vy - \mX\vbeta}^2,
$$

a scalar-valued function of the $2 \times 1$ vector $\vbeta$.
Its gradient is $-2\,\tp{\mX}(\vy - \mX\vbeta)$.
By @thm-first-order-condition, a minimizer of $f$ must make this gradient
$\vzero$, which is the equation $\tp{\mX}\mX\vbeta = \tp{\mX}\vy$.
The code below solves that equation and evaluates the gradient at
the solution and at $\vbeta = \vzero$.

::: {.panel-tabset}

#### Python

```{python}
#| code-fold: false
import numpy as np
import pandas as pd

auto = pd.read_csv("data/auto.csv")
y = auto["mpg"].to_numpy(dtype=float)
X = np.column_stack([np.ones(len(auto)), auto["horsepower"].to_numpy(dtype=float)])

def grad(beta):
    return -2 * X.T @ (y - X @ beta)

beta_hat = np.linalg.solve(X.T @ X, X.T @ y)
grad_hat = grad(beta_hat)
grad_zero = grad(np.zeros(2))

pd.DataFrame(
    {"beta": [beta_hat, np.zeros(2)], "gradient": [grad_hat, grad_zero]},
    index=["solution", "zero vector"],
)
```

A library fit gives the same coefficients:

```{python}
#| code-fold: false
import numpy as np

# np.polyfit returns the slope first, so reverse to (intercept, slope).
beta_polyfit = np.polyfit(auto["horsepower"], auto["mpg"], deg=1)[::-1]
beta_polyfit
```

#### R

```{r}
#| label: stationary-least-squares-r
#| code-fold: false
auto <- read.csv("data/auto.csv")
y <- auto$mpg
X <- cbind(1, auto$horsepower)

grad <- function(beta) -2 * t(X) %*% (y - X %*% beta)

beta_hat <- solve(t(X) %*% X, t(X) %*% y)
grad_hat <- grad(beta_hat)
grad_zero <- grad(c(0, 0))

list(
  solution = list(beta = c(beta_hat), gradient = c(grad_hat)),
  zero_vector = list(beta = c(0, 0), gradient = c(grad_zero))
)
```

A library fit gives the same coefficients:

```{r}
#| label: stationary-least-squares-lm
#| code-fold: false
coef(lm(mpg ~ horsepower, data = auto))
```

:::

```{r}
#| include: false
# The claims the prose makes: both languages agree, the gradient vanishes
# (relative to its size at the zero vector) at the solution, and the solution
# matches R's lm().
stopifnot(
  max(abs(c(beta_hat) - reticulate::py$beta_hat)) < 1e-6,
  max(abs(grad_hat)) < 1e-6 * max(abs(grad_zero)),
  max(abs(c(beta_hat) - coef(lm(mpg ~ horsepower, data = auto)))) < 1e-8,
  max(abs(c(beta_hat) - reticulate::py$beta_polyfit)) < 1e-6
)
```

The solution is an intercept of `{r} sprintf("%.1f", beta_hat[1])`
miles per gallon and a slope of `{r} sprintf("%.3f", beta_hat[2])`
miles per gallon per unit of horsepower.
The largest entry of the gradient there is
`{r} format(max(abs(grad_hat)), digits = 2)`,
zero up to rounding error,
against `{r} format(max(abs(grad_zero)), digits = 2, big.mark = ",")`
at the zero vector.
So the solution is a stationary point (@def-stationary-point),
and the zero vector is not.
The library fits above agree with the solution of that equation.

:::
