::::{#exm-ftc-cdf-pdf}
#### CDF and PDF of the exponential distribution

In what follows, $f$ denotes the PDF and $F$ the CDF --- the same letters
as the antiderivative pair in @def-antiderivative, because the FTC will
show $F$ is exactly an antiderivative of $f$.

For the exponential distribution with rate parameter $\lambda > 0$,
the probability density function (PDF) is
[@kleinbaum2012survival, Section II, p. 295, "Survival and Hazard
Functions for Selected Distributions"]:

$$f(t) = \lambda \ef{-\lambda t}, \quad t \ge 0$$

**FTC Part 2** gives the cumulative distribution function (CDF) from the PDF.
Apply the $\ef{cx}$ rule from @thm-integral-rules with $c = -\lambda$ to
antidifferentiate the integrand:

$$
\ba
F(t)
&= \int_0^t \lambda \ef{-\lambda u}\,du \\
&= \sb{\lambda \cd \frac{1}{-\lambda}\ef{-\lambda u}}_{u=0}^{u=t} \\
&= \sb{(-1)\ef{-\lambda u}}_{u=0}^{u=t} \\
&= \sb{-\ef{-\lambda u}}_{u=0}^{u=t} \\
&= -\ef{-\lambda t} - \paren{-\ef{0}} \\
&= -\ef{-\lambda t} - (-1) \\
&= 1 - \ef{-\lambda t}
\ea
$$

**FTC Part 1** recovers the PDF from the CDF:

$$
\ba
\deriv{t} F(t)
&= \deriv{t}\paren{1 - \ef{-\lambda t}} \\
&= 0 - (-\lambda)\ef{-\lambda t} \\
&= \lambda\ef{-\lambda t} \\
&= f(t)
\ea
$$

For a concrete instance: with $\lambda = 1$ (standard exponential),
the probability that $T \le 2$ is:

$$
F(2) = 1 - \ef{-1 \cd 2} = 1 - \ef{-2} \approx 1 - 0.135 = 0.865
$$

See @fig-exp-pdf-cdf.

```{r}
#| label: fig-exp-pdf-cdf-setup
#| include: false
lambda <- 1
t_focus <- 2
t_max <- 5
F_at_focus <- 1 - exp(-lambda * t_focus)
```

{{< slidebreak >}}

:::{#fig-exp-pdf-cdf layout-ncol="2"}

:::{#fig-exp-pdf-cdf-pdf}

```{r}
#| label: fig-exp-pdf-cdf-pdf-code
#| code-fold: true
ggplot2::ggplot() +
  ggplot2::geom_area(
    data = data.frame(t = seq(0, t_focus, length.out = 300)),
    ggplot2::aes(x = t, y = lambda * exp(-lambda * t)),
    fill = "steelblue", alpha = 0.4
  ) +
  ggplot2::geom_function(
    fun = \(t) lambda * exp(-lambda * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  ggplot2::labs(x = "t", y = "f(t)") +
  ggplot2::theme_minimal()
```

PDF with $\lambda = `r lambda`$; shaded area equals $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
:::

:::{#fig-exp-pdf-cdf-cdf}

```{r}
#| label: fig-exp-pdf-cdf-cdf-code
#| code-fold: true
ggplot2::ggplot() +
  ggplot2::geom_function(
    fun = \(t) 1 - exp(-lambda * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = t_focus, y = F_at_focus),
    size = 3, color = "steelblue"
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(x = t_focus, xend = t_focus, y = 0, yend = F_at_focus),
    linetype = "dashed", color = "steelblue"
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(x = 0, xend = t_focus, y = F_at_focus, yend = F_at_focus),
    linetype = "dashed", color = "steelblue"
  ) +
  ggplot2::labs(x = "t", y = "F(t)") +
  ggplot2::theme_minimal()
```

CDF with $\lambda = `r lambda`$; point marks $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
:::

Exponential distribution with $\lambda = `r lambda`$.
Left: the PDF $f(t) = \lambda \ef{-\lambda t}$; the shaded area under the curve from $0$ to $`r t_focus`$ equals $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
Right: the CDF $F(t) = 1 - \ef{-\lambda t}$; the dashed lines mark the value $F(`r t_focus`)$ computed via FTC Part 2.

:::

::::
