::::{#exm-ftc-part1-accumulation}
#### FTC Part 1 visualized: accumulation function for $f(t) = 2t$

Take $f(t) = 2t$ on $[0, 2]$. The accumulation function from $0$ is

$$F(x) \;\eqdef\; \int_0^x 2t\,dt \;=\; \sb{t^2}_{t=0}^{t=x} \;=\; x^2 - 0^2 \;=\; x^2,$$

so $F(x) = x^2$, and indeed $F'(x) = 2x = f(x)$, as @thm-ftc Part 1
predicts.
@fig-ftc-part1 shows the integrand on the left
(shaded area equals $F(x)$ at each $x$)
and the accumulation function $F(x) = x^2$ on the right
(its slope at $x$ equals $f(x) = 2x$).

{{< slidebreak >}}

```{r}
#| label: fig-ftc-part1-setup
#| include: false
x_marks <- c(1, 1.5, 2)
x_focus <- 1.5
```

:::{#fig-ftc-part1 layout-ncol="2"}

:::{#fig-ftc-part1-left}

```{r}
#| label: fig-ftc-part1-left-code
#| code-fold: true
ggplot2::ggplot() +
  ggplot2::geom_area(
    data = data.frame(t = seq(0, x_focus, length.out = 200)),
    ggplot2::aes(x = t, y = 2 * t),
    fill = "steelblue", alpha = 0.4
  ) +
  ggplot2::geom_function(fun = \(t) 2 * t, xlim = c(0, 2.2), linewidth = 1) +
  ggplot2::geom_vline(
    data = data.frame(x = x_marks),
    ggplot2::aes(xintercept = x, color = factor(x)),
    linetype = "dashed", linewidth = 0.6
  ) +
  ggplot2::labs(x = "t", y = "f(t) = 2t", color = "x") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

$f(t) = 2t$; shaded area equals $F(`r x_focus`) = `r x_focus^2`$; vertical lines mark $x \in \{`r paste(x_marks, collapse = ", ")`\}$.
:::

:::{#fig-ftc-part1-right}

```{r}
#| label: fig-ftc-part1-right-code
#| code-fold: true
slope_df <- data.frame(
  x = x_marks,
  Fx = x_marks^2,
  slope = 2 * x_marks
)

ggplot2::ggplot() +
  ggplot2::geom_function(fun = \(x) x^2, xlim = c(0, 2.2), linewidth = 1) +
  ggplot2::geom_point(
    data = slope_df,
    ggplot2::aes(x = x, y = Fx, color = factor(x)),
    size = 3
  ) +
  ggplot2::geom_segment(
    data = slope_df,
    ggplot2::aes(
      x = x - 0.3, xend = x + 0.3,
      y = Fx - 0.3 * slope, yend = Fx + 0.3 * slope,
      color = factor(x)
    ),
    linewidth = 0.8
  ) +
  ggplot2::labs(x = "x", y = expression(F(x) == x^2), color = "x") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

$F(x) = x^2$; tangent slope at each marked $x$ equals $f(x) = 2x$.
:::

Left: $f(t) = 2t$; the shaded area $\int_0^{`r x_focus`} 2t\,dt = F(`r x_focus`) = `r x_focus^2`$; vertical lines mark $x \in \{`r paste(x_marks, collapse = ", ")`\}$.
Right: $F(x) = x^2$; for each marked $x$, the tangent slope equals $f(x) = 2x$.

:::

::::
