::::{#exm-ftc-cdf-pdf}
#### CDF and PDF of the exponential distribution

In what follows, $f$ denotes the PDF and $F$ the CDF --- the same letters
as the antiderivative pair in @def-antiderivative, because the FTC will
show $F$ is exactly an antiderivative of $f$.

Let $T$ be a
[random variable](https://morrison-lab.github.io/pds/random-variables.html#def-random-variable)
with the [exponential distribution](https://morrison-lab.github.io/pds/random-variables.html#def-exponential)
with rate $\rate > 0$.
Its [probability density function (PDF)](https://morrison-lab.github.io/pds/random-variables.html#def-pdf) is
[@kleinbaum2012survival, Section II, p. 295, "Survival and Hazard
Functions for Selected Distributions"]:

$$f(t) = \rate \ef{-\rate t}, \quad t \ge 0$$

**FTC Part 2** gives the
[cumulative distribution function (CDF)](https://morrison-lab.github.io/pds/random-variables.html#def-cdf),
$F(t) = P(T \le t)$, from the PDF.
Apply the $\ef{cx}$ rule from @thm-integral-rules with $c = -\rate$ to
antidifferentiate the integrand:

$$
\ba
F(t)
&= \int_0^t \rate \ef{-\rate u}\,du
&& \text{(the CDF integrates the PDF)}
\\
&= \sb{\rate \cd \frac{1}{-\rate}\ef{-\rate u}}_{u=0}^{u=t}
&& \text{(FTC Part 2, with the } \ef{cx} \text{ rule)}
\\
&= \sb{(-1)\ef{-\rate u}}_{u=0}^{u=t}
&& \text{(} \rate / (-\rate) = -1 \text{)}
\\
&= \sb{-\ef{-\rate u}}_{u=0}^{u=t}
&& \text{(multiply by } -1 \text{)}
\\
&= -\ef{-\rate t} - \paren{-\ef{0}}
&& \text{(evaluate at the limits)}
\\
&= -\ef{-\rate t} - (-1)
&& \text{(} \ef{0} = 1 \text{)}
\\
&= 1 - \ef{-\rate t}
&& \text{(rearrange)}
\ea
$$

**FTC Part 1** recovers the PDF from the CDF:

$$
\ba
\deriv{t} F(t)
&= \deriv{t}\paren{1 - \ef{-\rate t}}
&& \text{(substitute } F \text{)}
\\
&= \deriv{t} 1 - \deriv{t} \ef{-\rate t}
&& \text{(derivative of a difference)}
\\
&= 0 - \deriv{t} \ef{-\rate t}
&& \text{(constant rule)}
\\
&= 0 - \ef{-\rate t} \cd \deriv{t}(-\rate t)
&& \text{(chain rule, with inner function } -\rate t \text{)}
\\
&= 0 - \ef{-\rate t} \cd (-\rate)
&& \text{(constant multiple rule)}
\\
&= \rate\ef{-\rate t}
&& \text{(simplify)}
\\
&= f(t)
&& \text{(definition of } f \text{)}
\ea
$$

For a concrete instance: with $\rate = 1$,
the probability that $T \le 2$ is:

$$
\begin{aligned}
F(2)
&= 1 - \ef{-1 \cd 2} \\
&= 1 - \ef{-2} \\
&\approx 1 - 0.135 \\
&= 0.865
\end{aligned}
$$

See @fig-exp-pdf-cdf.

```{r}
#| label: exp-pdf-cdf-setup
#| include: false
lambda <- 1
t_focus <- 2
t_max <- 5
F_at_focus <- 1 - exp(-lambda * t_focus)
```

{{< slidebreak >}}

:::{#fig-exp-pdf-cdf layout-ncol="2"}

:::{#fig-exp-pdf-cdf-pdf}

```{r}
#| label: exp-pdf-cdf-pdf-code
#| code-fold: true
ggplot2::ggplot() +
  ggplot2::geom_area(
    data = data.frame(t = seq(0, t_focus, length.out = 300)),
    ggplot2::aes(x = t, y = lambda * exp(-lambda * t)),
    fill = "steelblue", alpha = 0.4
  ) +
  ggplot2::geom_function(
    fun = \(t) lambda * exp(-lambda * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  ggplot2::labs(x = "t", y = "f(t)") +
  ggplot2::theme_minimal()
```

PDF with $\rate = `r lambda`$; shaded area equals $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
:::

:::{#fig-exp-pdf-cdf-cdf}

```{r}
#| label: exp-pdf-cdf-cdf-code
#| code-fold: true
ggplot2::ggplot() +
  ggplot2::geom_function(
    fun = \(t) 1 - exp(-lambda * t),
    xlim = c(0, t_max), linewidth = 1
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = t_focus, y = F_at_focus),
    size = 3, color = "steelblue"
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(x = t_focus, xend = t_focus, y = 0, yend = F_at_focus),
    linetype = "dashed", color = "steelblue"
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(x = 0, xend = t_focus, y = F_at_focus, yend = F_at_focus),
    linetype = "dashed", color = "steelblue"
  ) +
  ggplot2::labs(x = "t", y = "F(t)") +
  ggplot2::theme_minimal()
```

CDF with $\rate = `r lambda`$; point marks $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
:::

Exponential distribution with $\rate = `r lambda`$.
Left: the PDF $f(t) = \rate \ef{-\rate t}$; the shaded area under the curve from $0$ to $`r t_focus`$ equals $F(`r t_focus`) \approx `r round(F_at_focus, 3)`$.
Right: the CDF $F(t) = 1 - \ef{-\rate t}$; the dashed lines mark the value $F(`r t_focus`)$ computed via FTC Part 2.

:::

::::
