:::{#exm-function-profile-logistic}
#### The profile of the logistic function

The **logistic function** (also called $\expit$ or the sigmoid) is

$$
f(x) = \frac{1}{1 + e^{-x}}.
$$

The profile in @def-function-profile of $f$ is as follows.
Write $E = e^{-x}$, so that $f = \frac{1}{1 + E}$ and $1 - f = \frac{E}{1 + E}$.

- **Logarithm:** $\logf{f(x)} = -\logf{1 + e^{-x}}$, defined for every real $x$ because $f(x) > 0$.
- **Derivative:** by the chain rule,

  $$
  \begin{aligned}
  f'(x)
  &= \frac{e^{-x}}{\paren{1 + e^{-x}}^2}
  && \text{(chain rule applied to } (1 + e^{-x})^{-1} \text{)} \\
  &= \frac{1}{1 + E} \cdot \frac{E}{1 + E}
  && \text{(split the square)} \\
  &= f(x)\,\paren{1 - f(x)}.
  && \text{(identify } f \text{ and } 1 - f \text{)}
  \end{aligned}
  $$

  Differentiating again with the product rule gives
  $f''(x) = f'(1 - f) - f f' = f(x)\paren{1 - f(x)}\paren{1 - 2f(x)}$.
- **Logarithmic derivative:** $\frac{f'(x)}{f(x)} = 1 - f(x)$,
  so the relative rate of change of $f$ is the complement of $f$ itself.
- **Inverse:** for $y \in (0, 1)$, solving $y = \frac{1}{1 + e^{-x}}$ for $x$ gives

  $$
  \begin{aligned}
  \frac{1}{y} &= 1 + e^{-x} \\
  e^{-x} &= \frac{1 - y}{y} \\
  x &= \logf{\frac{y}{1 - y}}.
  \end{aligned}
  $$

  So $f^{-1}(y) = \logitf{y}$, the log-odds.
  The inverse exists because $f' = f(1 - f) > 0$, so $f$ is strictly increasing and therefore one-to-one.
- **Complement:** $1 - f(x) = \frac{e^{-x}}{1 + e^{-x}} = \frac{1}{e^{x} + 1} = f(-x)$.
- **Reciprocal:** $\frac{1}{f(x)} = 1 + e^{-x}$.
- **Antiderivative:** $F(x) = \logf{1 + e^{x}}$ satisfies $F'(x) = \frac{e^{x}}{1 + e^{x}} = f(x)$,
  so $\int f(x)\,dx = \logf{1 + e^{x}} + C$.
- **Key values:** $f(0) = \frac{1}{1 + 1} = \frac{1}{2}$ and $f(1) = \frac{1}{1 + e^{-1}}$.
  As $x \to \infty$, $e^{-x} \to 0$, so $f(x) \to 1$;
  as $x \to -\infty$, $e^{-x} \to \infty$, so $f(x) \to 0$.

The code checks these results numerically,
using the base R functions `plogis()` for $f$ and `qlogis()` for $f^{-1}$,
and `D()` to differentiate symbolically.

```{r}
#| label: function-profile-logistic
#| code-fold: false
f <- plogis
x <- c(-2, -0.5, 0, 1, 3)

f_expr <- quote(1 / (1 + exp(-x)))
f_prime <- eval(D(f_expr, "x"), list(x = x))
f_second <- eval(D(D(f_expr, "x"), "x"), list(x = x))

checks <- c(
  derivative = max(abs(f_prime - f(x) * (1 - f(x)))),
  second_derivative = max(abs(f_second - f(x) * (1 - f(x)) * (1 - 2 * f(x)))),
  log_derivative = max(abs(f_prime / f(x) - (1 - f(x)))),
  inverse = max(abs(qlogis(f(x)) - x)),
  complement = max(abs(1 - f(x) - f(-x))),
  reciprocal = max(abs(1 / f(x) - (1 + exp(-x)))),
  antiderivative = abs(
    integrate(f, 0, 1)$value - (log(1 + exp(1)) - log(2))
  )
)
signif(checks, 3)
stopifnot(all(checks < 1e-14))

key_values <- c(f(0), f(1), f(10), f(-10))
```

Each entry of `checks` is the largest discrepancy between a formula above and a numerical computation;
every one is below $10^{-14}$ (the chunk stops with an error otherwise), which is rounding error.
The key values are $f(0) =$ `{r} key_values[1]` and $f(1) \approx$ `{r} round(key_values[2], 4)`.
At $x = 10$ and $x = -10$ the function is already close to its limits,
$f(10) \approx$ `{r} round(key_values[3], 5)` and $f(-10) \approx$ `{r} format(key_values[4], digits = 3, scientific = FALSE)`.
@fig-function-profile-logistic graphs $f$, $f'$ and $\logf{f}$.

```{r}
#| label: fig-function-profile-logistic
#| fig-cap: >-
#|   The logistic function, its derivative and its logarithm
#| fig-alt: >-
#|   Three panels over x from -6 to 6.
#|   The first, the logistic function, rises in an S shape from near 0 to near 1
#|   and passes through 0.5 at x = 0.
#|   The second, its derivative, is a bell-shaped bump
#|   with maximum 0.25 at x = 0 and tails near 0.
#|   The third, the logarithm of the logistic function,
#|   increases from a steep line of slope 1 on the left
#|   and flattens toward 0 on the right.
grid <- seq(-6, 6, length.out = 241)
panels <- data.frame(
  x = rep(grid, 3),
  panel = factor(
    rep(c("f(x)", "f'(x)", "log f(x)"), each = length(grid)),
    levels = c("f(x)", "f'(x)", "log f(x)")
  ),
  value = c(f(grid), f(grid) * (1 - f(grid)), log(f(grid)))
)
ggplot2::ggplot(panels) +
  ggplot2::aes(x = x, y = value) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~panel, scales = "free_y") +
  ggplot2::labs(x = "x", y = NULL)
```

:::
