::::{#exm-fubini-fail}

#### When conditions fail: a counterexample

The conditions in @thm-fubini are not merely technical ---
when they fail, iterated integrals can exist yet disagree.

Let
$$f(x, y) = \frac{x^2 - y^2}{(x^2 + y^2)^2}$$
on the unit square $R = [0, 1] \times [0, 1]$.
Strictly, $f$ is defined on $R \setminus \{(0, 0)\}$:
the denominator vanishes at the origin, so $f$ is undefined there
(we return to this point in the condition check).

**Integrating $y$ first, then $x$:**

Holding $x$ fixed and differentiating in $y$
(by the quotient rule, @thm-quotient-rule),

$$
\begin{aligned}
\dif{}{y}\frac{y}{x^2 + y^2} &= \frac{(x^2 + y^2) - y \cd 2y}{(x^2 + y^2)^2} \\
&= \frac{x^2 - y^2}{(x^2 + y^2)^2}.
\end{aligned}
$$

(A derivative in one variable with the others held fixed is a
[partial derivative](vector-calculus.qmd#def-partial-derivative),
defined on the vector calculus page.)
The arctangent $\arctan$ is the inverse of the tangent function from trigonometry;
all this example needs is that $\dif{}{x}\arctan(x) = \frac{1}{1 + x^2}$,
$\arctan(0) = 0$, and $\arctan(1) = \frac{\pi}{4}$.

$$
\ba
\int_0^1\!\int_0^1 f(x, y)\,dy\,dx
&= \int_0^1 \sb{\frac{y}{x^2 + y^2}}_{y=0}^{y=1}\,dx
\\&= \int_0^1 \frac{1}{x^2 + 1}\,dx
\\&= \sb{\arctan(x)}_0^1
\\&= \frac{\pi}{4}
\ea
$$

**Integrating $x$ first, then $y$:**

Holding $y$ fixed and differentiating in $x$,

$$
\begin{aligned}
\dif{}{x}\paren{-\frac{x}{x^2 + y^2}} &= -\frac{(x^2 + y^2) - x \cd 2x}{(x^2 + y^2)^2} \\
&= \frac{x^2 - y^2}{(x^2 + y^2)^2}:
\end{aligned}
$$

$$
\ba
\int_0^1\!\int_0^1 f(x, y)\,dx\,dy
&= \int_0^1 \sb{-\frac{x}{x^2 + y^2}}_{x=0}^{x=1}\,dy
\\&= \int_0^1 \paren{-\frac{1}{1 + y^2}}\,dy
\\&= -\sb{\arctan(y)}_0^1
\\&= -\frac{\pi}{4}
\ea
$$

**Conclusion:** $\dfrac{\pi}{4} \neq -\dfrac{\pi}{4}$,
so the two iterated integrals are unequal.
@thm-fubini does not apply here.

**Why @thm-fubini's condition fails:**
@thm-fubini requires $f$ to be **continuous** on $R$.
The denominator $(x^2 + y^2)^2$ vanishes at the origin $(0, 0) \in R$,
so $f$ is *not even defined* there --- let alone continuous --- and the
theorem does not apply.

[@wp:fubini]

@fig-fubini-fail shows the surface.
The failure comes from the origin, where $f$ is undefined and unbounded.

:::{#fig-fubini-fail}

```{r}
#| code-fold: true
#| message: false
#| echo: !expr knitr::is_html_output()
#| eval: !expr knitr::is_html_output()

n_grid <- 81
eps <- 0.04
x_seq <- seq(eps, 1, length.out = n_grid)
y_seq <- seq(eps, 1, length.out = n_grid)

z_mat <- outer(x_seq, y_seq, function(x, y) (x^2 - y^2) / (x^2 + y^2)^2)
z_clip <- 50
z_mat[z_mat > z_clip] <- z_clip
z_mat[z_mat < -z_clip] <- -z_clip

plotly::plot_ly(x = ~x_seq, y = ~y_seq, z = ~t(z_mat)) |>
  plotly::add_surface(
    showscale = FALSE,
    colorscale = list(
      list(0, "#3b4cc0"),
      list(0.5, "#dddddd"),
      list(1, "#b40426")
    )
  ) |>
  plotly::layout(scene = list(
    xaxis = list(title = "x"),
    yaxis = list(title = "y"),
    zaxis = list(title = "f(x, y)", range = c(-z_clip, z_clip)),
    camera = list(eye = list(x = 1.6, y = 1.6, z = 0.6))
  ))
```

```{r}
#| echo: false
#| eval: !expr isFALSE(knitr::is_html_output())
#| results: asis

cat(
  "*(This 3D surface is interactive; see the HTML or slide version",
  "of this page to view it.)*"
)
```

Surface $f(x, y) = (x^2 - y^2)/(x^2 + y^2)^2$ on $[0, 1]^2$,
sampled away from the origin and clipped to $[-50, 50]$ for display.
Approaching the origin, the function grows without bound along the $x$-axis (red ridge,
$f > 0$ when $|x| > |y|$) and falls without bound along the $y$-axis (blue
ridge, $f < 0$ when $|y| > |x|$). Because $f$ is undefined at $(0, 0)$,
$f$ is not continuous on $R$ and @thm-fubini does not apply.

:::

::::
